---
title: "Note Frequencies"
author: 
 - name: Thomas Sostarics
 - url: https://tsostarics.com/
 - affiliation: Northwestern University Department of Linguistics
 - orcid: 0000-0002-1178-7967
date: '2022-07-13'
editor: source
description: "A quick reference for musical notes and their corresponding frequencies and wavelengths."
categories: ['Reference']
code-tools: false
code-fold: 'show'
execute:
  message: false
  echo: false
  warning: false
---

I've had Bryan Suits' website bookmarked for years specifically for the page on converting musical notes to frequencies and wavelengths.
I have learned that he has [sadly passed away](https://blogs.mtu.edu/physics/2023/10/professor-emeritus-bryan-suits-passes-away/) recently, and Michigan Tech removed his website in the process.
This is a recreation of the table on his website, mostly so I can access it again more easily.

## Frequencies for equal-tempered scale

```{r}
#| echo: false
library(sosprosody)
library(gt)
note_names <- 
  c("C0","C#0/Db0","D0","D#0/Eb0","E0","F0","F#0/Gb0","G0","G#0/Ab0","A0","A#0/Bb0","B0")

note_names <- 
  c(lapply(0:8, \(octave) gsub("0", octave, note_names)), recursive = TRUE)

semitones <- -57:50

make_table <- function(a4, semitones) {
  note_freqs <- add_semitones(a4, semitones)
  data.frame(Note = note_names,
             Frequency = round(note_freqs, 2),
             Wavelength = round(hz_to_wavelength(note_freqs, "cm"),2)) |> 
    gt::gt() |> 
    gt::cols_label(Frequency= "Frequency (Hz)", Wavelength = "Wavelength (cm)")
}

```


Speed of Sound = 345 m/s = 1130 ft/s = 770 miles/hr.
To convert lengths in cm to inches, divide by 2.54.


Select a frequency for A4 below. "Middle C" is C4.


::: {.panel-tabset}


## **440**

```{r}
make_table(440, semitones)

```

## 432

```{r}
make_table(432, semitones)
```


## 434

```{r}
make_table(434, semitones)

```

## 436

```{r}
make_table(436, semitones)

```

## 438

```{r}
make_table(438, semitones)

```




## 442

```{r}
make_table(442, semitones)

```


## 444

```{r}
make_table(444, semitones)

```


## 446

```{r}
make_table(446, semitones)

```

:::

Wayback machine links below:

 - [Scales: Just vs Equal Temperament    (and related topics)](https://web.archive.org/web/20180205170544/http://pages.mtu.edu/~suits/scales.html)
 - [Equations](https://web.archive.org/web/20180118163813/http://pages.mtu.edu/~suits/NoteFreqCalcs.html)
 - [Note Frequency Table](https://web.archive.org/web/20180202150802/http://pages.mtu.edu/~suits/notefreqs.html)
 - [More about Speed of Sound.](https://web.archive.org/web/20180131041533/http://pages.mtu.edu/~suits/SpeedofSound.html)



```{r}
sessionInfo()
```

